---
title: 'Analytics Edge: Unit 1'
author: "Sulman Khan"
date: "October 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction to Analytics Edge

### Prevalence of Data

* 2.7 Zettabytes of  electronic data exist in the world 
today - 2,700,000,000,000,000,000,000 bytes 
    + This is equal to the storage required for more than 200 
billion HD movies
* New data is produced at an exponential rate
* Decoding the human genome originally took 10 years 
to process; now it can be achieved in one week 

### Data and Analytics are Useful

* Estimated that there is a shortage of  140,000 - 190,000 
people with deep analytical skills to fill the demand of  
jobs in the U.S. by 2018 
* IBM has invested over $20 billion since 2005 to grow its 
analytics business 
* Companies will invest more than $120 billion by 2015 on 
analytics, hardware, software and services 
* Critical in almost every industry 
    + Healthcare, media, sports, finance, government, etc.
    
### Definition of Analytics
The science of using **data** to build **models** that lead to better **decisions** that add **value** to individuals, to companies, and to institutions.


### Examples of Data Anlytics Used
#### IBM Watson

* Watson is a supercomputer with 3,000 processors 
and a database of 200 million pages of information 
* Watson combined many algorithms to increase 
accuracy and confidence 
* Approached the problem in a different way than how 
a human does 
* Deals with massive amounts of  data, often in 
unstructured form 
    + 90% of data in the world is unstructured 


#### eHarmony
* Online dating site focused on long term relationships 
* Relies much more on data than other dating sites 
* Suggests a limited number of high quality matches
    + Users don't have to search and dig through profiles 
* eHarmony has successfully leveraged the power of  
analytics to create a successful and thriving business 
    + 14% of US online dating market 
    
#### The Framingham Heart Study
* Much of the now-common knowledge regarding heart disease came from this study
* Provided necessary evidence for the development of drugs to lower blood pressure 
* Paved the way for other clinical prediction rules
    + Predict clinical outcomes using patient data
* A model allows medical professionals to make predictions for patients worldwide 

#### D2Hawkeye
* Combined data with analytics to improve quality and 
cost management in healthcare 
* Substantial improvement in D2Hawkeye's ability to 
identify patients who need more attention 
* Use expert knowledge to identify new variables and 
refine existing variables 
* Can make predictions for millions of  patients 
without manually reading patient files 

## An Introduction to R

### What is R?
* A software environment for data analysis, statistical computing, and graphics 
* A programming language 

In the next section, the basic operations and functions used in R for data analysis are explored.

### Basic Calculations
```{r Basic Calculations}
8*6
2^16
8*6
8*10
```

### Functions
``` {r Functions}
sqrt(2)
abs(-65)
```

### Variables
``` {r Variables}
SquareRoot2 = sqrt(2)
SquareRoot2
HoursYear <- 365*24
HoursYear
ls()
```

### Vectors
Two vectors - **Country** and **LifeExpectancy** are created. Accordingly, both vectors are indexed to display specific elements inside the vectors. Finally, a third vector **Sequence** has a range from 0 to 100 in increments of 2.
``` {r Vectors}
c(2,3,5,8,13)
Country = c("Brazil", "China", "India","Switzerland","USA")
LifeExpectancy = c(74,76,65,83,79)
Country
LifeExpectancy
Country[1]
LifeExpectancy[3]
Sequence = seq(0,100,2)
Sequence
```
### Data Frames
The data.frame **CountryData** calls upon the two vectors **Country** and **LifeExpectancy**. Successivly, an additional vector **CountryData$Populations** is added to the data.frame. Finally, a data.frame **AllCountryData** utilizes the two previous data.frames **CountryData** and **NewCountryData**.
``` {r Data Frames}
CountryData = data.frame(Country, LifeExpectancy)
CountryData
CountryData$Population = c(199000,1390000,1240000,7997,318000)
CountryData
Country = c("Australia","Greece")
LifeExpectancy = c(82,81)
Population = c(23050,11125)
NewCountryData = data.frame(Country, LifeExpectancy, Population)
NewCountryData
AllCountryData = rbind(CountryData, NewCountryData)
AllCountryData
```

### Loading csv files
The dataset **WHO.csv** is loaded into the variable **WHO**. The str and summary commands provide physical and statistical descriptions of the variable. 
``` {r Loading csv files}
WHO = read.csv("WHO.csv")
str(WHO)
summary(WHO)
```
### Subsetting
The dataset **WHO** is subsetted into **WHO_Europe** using the region as an argument to collect the data.
``` {r Subsetting}
WHO_Europe = subset(WHO, Region == "Europe")
str(WHO_Europe)
```
### Writing csv files
``` {r Writing csv files}
write.csv(WHO_Europe, "WHO_Europe.csv")
```
### Removing variables
``` {r Removing variables}
rm(WHO_Europe)
```

### Basic Data Analysis
Various basic data analysis commands are implemented to provide a statistical summary of the datasets and how to appropriately identify various elements of interest.
``` {r Basic Data Analysis}
mean(WHO$Under15)
sd(WHO$Under15)
summary(WHO$Under15)

which.min(WHO$Under15)
WHO$Country[86]

which.max(WHO$Under15)
WHO$Country[124]
```

### Scatterplot
``` {r Scatterplot}
plot(WHO$GNI, WHO$FertilityRate)
```

### Subsetting Outliers
``` {r Subsetting Outliers}
Outliers = subset(WHO, GNI > 10000 & FertilityRate > 2.5) 
nrow(Outliers)
Outliers[c("Country","GNI","FertilityRate")]
```   

### Histograms
``` {r Histogram}
hist(WHO$CellularSubscribers)
```
### Boxplot
``` {r Boxplot}
boxplot(WHO$LifeExpectancy ~ WHO$Region)
boxplot(WHO$LifeExpectancy ~ WHO$Region, xlab = "", ylab = "Life Expectancy", main = "Life Expectancy of Countries by Region")
```

### Summary Tables
The **table** command creates a table for each region in the **WHO** data.frame.
The **tapply** command demonstrates the relationship between two vectors in the data.frame using a statistical descriptor.
``` {r Summary Tables}
table(WHO$Region)

tapply(WHO$Over60, WHO$Region, mean)
tapply(WHO$LiteracyRate, WHO$Region, min)
tapply(WHO$LiteracyRate, WHO$Region, min, na.rm=TRUE)
```

## Understanding Food USDA Dataset
In the following section, the United States Department of Agriculture (USDA) dataset on the dietary contents of food is examined.


### Loading in the Dataset
#### Read the csv file
``` {r USDA_read}
  USDA = read.csv("USDA.csv")
```

#### Structure of the dataset
``` {r USDA_str}
  str(USDA)
```

#### Statistical summary
``` {r USDA_summary}
  summary(USDA)
````

### Basic Data Analysis

#### Vector notation
``` {r USDA_vector_notation, results = FALSE }
  USDA$Sodium
```

#### Finding the index of the food with highest sodium levels
``` {r USDA_indexSodium}
  which.max(USDA$Sodium)
```

#### Get names of variables in the dataset
``` {r USDA_names}
  names(USDA)
```

#### Get the name of the food with highest sodium levels
``` {r USDA_HighestSodiumDescription}
  USDA$Description[265]
```

#### Create a subset of the foods with sodium content above 10,000mg
``` {r USDA_HighSodiumSubset}
  HighSodium = subset(USDA, Sodium>10000)
```

#### Count the number of rows, or observations
``` {r USDA_nrowsHighSodium}
nrow(HighSodium)
```

#### Output names of the foods with high sodium content
``` {r USDA_outputHighSodium}
  HighSodium$Description
```

#### Finding the index of CAVIAR in the dataset
``` {r USDA_CaviarIndex}
  match("CAVIAR", USDA$Description)
```

#### Find amount of sodium in caviar
``` {r USDA_indexSodium4154}
  USDA$Sodium[4154]
```

#### Doing it in one command!
``` {r USDA_CaviarIndexSodium}
  USDA$Sodium[match("CAVIAR", USDA$Description)]
```

#### Summary function over Sodium vector
``` {r USDA_SodiumSummary}
  summary(USDA$Sodium)
```

#### Standard deviation
``` {r USDA_SDSodium}
  sd(USDA$Sodium, na.rm = TRUE)
```

## Plots of USDA Dataset
### Scatter Plots

``` {r USDA_ScatterPlotbase}
  plot(USDA$Protein, USDA$TotalFat)
```

#### Add xlabel, ylabel and title
``` {r USDA_ScatterPlotlabels}
  plot(USDA$Protein, USDA$TotalFat, xlab="Protein", ylab = "Fat", main = "Protein vs Fat", col = "red")
````

### Histograms
``` {r USDA_Histogrambase}
  hist(USDA$VitaminC, xlab = "Vitamin C (mg)", main = "Histogram of Vitamin C")
```

#### Add limits to x-axis
``` {r USDA_Histogramxlimits}
  hist(USDA$VitaminC, xlab = "Vitamin C (mg)", main = "Histogram of Vitamin C", xlim = c(0,100))
```

#### Specify breaks of histogram
``` {r USDA_Histogrambreaks}
  hist(USDA$VitaminC, xlab = "Vitamin C (mg)", main = "Histogram of Vitamin C", xlim = c(0,100), breaks=100)
  hist(USDA$VitaminC, xlab = "Vitamin C (mg)", main = "Histogram of Vitamin C", xlim = c(0,100), breaks=2000)
```

### Boxplots
``` {r USDA_boxplotSugar}
  boxplot(USDA$Sugar, ylab = "Sugar (g)", main = "Boxplot of Sugar")
```

## Adding a variable

### Creating a variable that takes value 1 if the food has higher sodium than average, 0 otherwise
``` {r USDA_creatingvar}
  HighSodium = as.numeric(USDA$Sodium > mean(USDA$Sodium, na.rm=TRUE))
  str(HighSodium)
```  
### Adding the variable to the dataset
``` {r USDA_addingvariabletothedataset}
  USDA$HighSodium = as.numeric(USDA$Sodium > mean(USDA$Sodium, na.rm=TRUE))
```

### Similarly for HighProtein, HigCarbs, HighFat
``` {r USDA_AdditionalvariablesforPCF}
  USDA$HighCarbs = as.numeric(USDA$Carbohydrate > mean(USDA$Carbohydrate, na.rm=TRUE))
  USDA$HighProtein = as.numeric(USDA$Protein > mean(USDA$Protein, na.rm=TRUE))
  USDA$HighFat = as.numeric(USDA$TotalFat > mean(USDA$TotalFat, na.rm=TRUE))
```

## Summary Tables

### How many foods have higher sodium level than average?
``` {r USDA_tableHS}
  table(USDA$HighSodium)
```

### How many foods have both high sodium and high fat?
``` {r USDA_tableHS&HF}
  table(USDA$HighSodium, USDA$HighFat)
```

### Average amount of iron sorted by high and low protein?
``` {r USDA_tapplyIHPmean}
  tapply(USDA$Iron, USDA$HighProtein, mean, na.rm=TRUE)
```

### Maximum level of Vitamin C in hfoods with high and low carbs?
``` {r USDA_tapplyVCHCmax}
  tapply(USDA$VitaminC, USDA$HighCarbs, max, na.rm=TRUE)
```

### Using summary function with tapply
``` {r USDA_tapplyVCHCsummary}
  tapply(USDA$VitaminC, USDA$HighCarbs, summary, na.rm=TRUE)
```
