---
title: 'Analytics Edge: Unit 2 - NBA'
author: "Sulman Khan"
date: "October 26, 2018"
output: 
  html_document:
    css: C:/Users/Sulman/Desktop/analytics edge/gf_small_touches.css
    highlight: tango
    mathjax: default
    theme: cerulean
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
require(knitr)
options(width = 160, scipen = 5)
options(dplyr.print_max = 200)
# options(width = 100, digits = 7)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE, 
               collapse = TRUE, tidy = FALSE,
               cache = TRUE, cache.path = '.cache/', 
               fig.align = 'left', dpi = 100, fig.path = 'figures/Introductiontotextanalytics/')
# opts_chunk$set(dev="png", 
#                dev.args=list(type="cairo"),
#                dpi=96)
```

## NBA in R

### Linear Regression

#### Read in the data
```{r nbaread}
NBA = read.csv("NBA_train.csv")
str(NBA)
```

#### How many wins to make the playoffs?
```{r howmanywins}
table(NBA$W, NBA$Playoffs)
```

#### Compute Points Difference
```{r computerpointdiff}
NBA$PTSdiff = NBA$PTS - NBA$oppPTS
```

#### Check for linear relationship
```{r checlinrelationship}
plot(NBA$PTSdiff, NBA$W)
```

#### Linear regression model for wins
```{r linearregmodelwins}
WinsReg = lm(W ~ PTSdiff, data=NBA)
summary(WinsReg)
```

#### Linear regression model for points scored
```{r linearregpointsscored}
PointsReg = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + DRB + TOV + STL + BLK, data=NBA)
summary(PointsReg)
```

#### Sum of Squared Errors
```{r SSE}
PointsReg$residuals
SSE = sum(PointsReg$residuals^2)
SSE
```

#### Root mean squared error
```{r RMSE}
RMSE = sqrt(SSE/nrow(NBA))
RMSE
```

#### Average number of points in a season
```{r averagenumberofpoints}
mean(NBA$PTS)
```

#### Remove insignifcant variables
```{r removeinsig}
summary(PointsReg)

PointsReg2 = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + DRB + STL + BLK, data=NBA)
summary(PointsReg2)

PointsReg3 = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + STL + BLK, data=NBA)
summary(PointsReg3)

PointsReg4 = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + STL, data=NBA)
summary(PointsReg4)
```

#### Compute SSE and RMSE for new model
```{r computessermse}
SSE_4 = sum(PointsReg4$residuals^2)
RMSE_4 = sqrt(SSE_4/nrow(NBA))
SSE_4
RMSE_4
```


#### Read in test set
```{r readtestset}
NBA_test = read.csv("NBA_test.csv")
```

#### Make predictions on test set
```{r makepredtest}
PointsPredictions = predict(PointsReg4, newdata=NBA_test)
```

#### Compute out-of-sample R^2
```{r oosrsquared}
SSE = sum((PointsPredictions - NBA_test$PTS)^2)
SST = sum((mean(NBA$PTS) - NBA_test$PTS)^2)
R2 = 1 - SSE/SST
R2
```

#### Compute the RMSE
```{r computermsetest}
RMSE = sqrt(SSE/nrow(NBA_test))
RMSE
```